---
title: 'HW 5: Correlations and linear models. Tests for categorial variables'
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(vcd) # for mosaic plots
```

## 1. Vowel reduction in Russian
Pavel Duryagin ran an experiment on perception of vowel reduction in Russian language. The dataset `shva` includes the following variables:  
_time1_ - reaction time 1  
_duration_ - duration of the vowel in the stimuly (in milliseconds, ms)  
_time2_ - reaction time 2  
_f1_, _f2_, _f3_ - the 1st, 2nd and 3rd formant of the vowel measured in Hz (for a short introduction into formants, see [here](https://home.cc.umanitoba.ca/~krussll/phonetics/acoustic/formants.html))  
_vowel_ - vowel classified according the 3-fold classification (_A_ - _a_ under stress, _a_ - _a/o_ as in the first syllable before the stressed one, _y_ (stands for shva) - _a/o_ as in the second etc. syllable before the stressed one or after the stressed syllable, cf. _g_[_y_]_g_[_a_]_t_[_A_]_l_[_y_] _gogotala_ `guffawed').  
In this part, we will ask you to analyse correlation between f1, f2, and duration.
The dataset is available [https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt](here).

### 1.0 
Read the data from file to the variable `shva`.
```{r 1.0}
shva <- read_tsv("https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt")
shva
```

### 1.1 
Scatterplot `f1` and `f2` using `ggplot()`. 
Design it to look like the [following](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin1.png).
```{r 1.1}
ggplot(shva, aes(x=f2, y=f1, color=vowel)) + geom_point(shape=16) +
+     scale_x_reverse() + scale_y_reverse() + theme(legend.position="none")
```

### 1.2 
Plot the boxplots of `f1` and `f2` for each vowel using `ggplot()`.
Design it to look like [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin2.png) and [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin3.png).
```{r 1.2}
# f1 boxplot
 ggplot(shva, aes(x=vowel, y=f1, fill=vowel)) + 
+     geom_boxplot() + coord_flip() + theme(legend.position="none")
# f2 boxplot
 ggplot(shva, aes(x=vowel, y=f2, fill=vowel)) + 
+     geom_boxplot() + coord_flip() + theme(legend.position="none")
```

### 1.3 
Calculate Pearson's correlation of `f1` and `f2` (all data)
```{r 1.3}
cor(shva$f1, shva$f2, method = 'pearson')
```

### 
1.4 Calculate Pearson's correlation of `f1` and `f2` for each vowel
```{r 1.4}
shva %>% filter(vowel == 'y') -> vowel_y
cor(vowel_y$f1, vowel_y$f2, method = 'pearson')
shva %>% filter(vowel == 'A') -> vowel_A
cor(vowel_A$f1, vowel_A$f2, method = 'pearson')
shva %>% filter(vowel == 'a') -> vowel_a
cor(vowel_a$f1, vowel_a$f2, method = 'pearson')

```

## 2 Linear regressions

### 2.1.1 
Use the linear regression model to predict `f2` by `f1`.
```{r 2.1.1}
model <- lm(f2~f1, data=shva)
shva$model <- predict(model)
```

### 2.1.2 
Write down the equation for f2 using coefficients from the model (e.g. $y =  b + kx$)
```{r 2.1.2}
model
```

```
Call:
lm(formula = f2 ~ f1, data = shva)

Coefficients:
(Intercept)           f1  
  1639.7021      -0.4287  

f2 = 1639.7021 - 0.4287 * f1
```

### 2.1.3 
Provide the adjusted R$^2$
```{r 2.1.3}
summary(model)$adj.r.sq
```

```
0.3319054
```

### 2.1.4 
Add the regression line in the scatterplot 1.1.
```{r 2.1.4}
ggplot(shva, aes(x=f2, y=f1, color=vowel)) + geom_point(shape=16) + geom_smooth(method=lm, se = FALSE) + scale_x_reverse() + scale_y_reverse() + theme(legend.position="none") 
```

### 2.1.5 
Make a scatter plot for `f1` and `f2` grouped by vowels. 
Use `ggplot()` and `facet_wrap()`.

```{r 2.1.5}
ggplot(shva, aes(x=f2, y=f1, color=vowel)) + geom_point(shape=16) + geom_smooth(method=lm, se = FALSE) + scale_x_reverse() + scale_y_reverse() + facet_wrap( ~ vowel, ncol=2)
```

### 2.2.1 
Use the linear regression model to predict `f2` by `f1` and `vowel`.
```{r 2.2.1}
model2 <- lm(f2~f1+vowel, data = shva)
shva$model2 <- predict(model2)
```

### 2.2.2 
What is the intercept of the model?
```{r 2.2.2}
model2$coefficients[1]
```

### 2.2.3 
Provide the adjusted R$^2$
```{r 2.2.3}
summary(model2)$adj.r.sq
```
```
0.4801 
```
### 2.2.4 
Write down your general conclusions about the relationship between `f1`, `f2`, and `vowels`.

```
Considering both models, we see that `f1`doesn't influence the prediction of a variable `f2` so much (the signific. code show us). As for the second model, it performs a little bit better than the first one (as R square increases). So, it means that a variable `vowel` influences the prediction, and there is a association between `f2` and this one.  
```

## 3. Dutch causative constructions

When the Dutch use two near-synonymous periphrastic causative verbs, *doen* and *laten*?
```
       De politie deed/liet de auto stoppen.
  lit. the police did/let the car stop 
       'The police stopped the car'
```

This is a data set on two rival constructions with *doen* and *laten* sampled from the newspaper corpora. The data frame includes 500 observations on the following 7 variables:

* `Aux` -- verb: a factor with levels `doen` and `laten`  
* `CrSem` -- the semantic class of the Causer: a factor with levels `Anim` (animate) and `Inanim` (inanimate)  
* `CeSem` -- the semantic class of the Causee: a factor with levels `Anim` (animate) and `Inanim` (inanimate)  
* `CdEvSem` -- the semantic domain of the caused event expressed by the Effected Predicate: a factor with levels `Ment` (mental) and `NonMent` (e.g. physical or social)  
* `CeSynt` -- the syntactic status of the Causee: a factor with levels `Clause`, `Impl` (implicit, not expressed), `NP` (noun phrase), `PP` (prepositional phrase)    
* `EPTrans` -- transitivity or intransitivity of the effected predicate, a factor with two levels `Tr` and `Intr`  
* `Country` -- a factor with levels `BE` (Belgium) and `NL` (Netherlands)  
* `Domain` -- a factor with four levels for newspaper domains.    

Data from Natalya Levshina's `RLing` package available (here)[https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv]
Read more on the constructions in [Levhina, Geerarts, Speelman 2014](https://www.academia.edu/7287585/Dutch_causative_constructions_Quantification_of_meaning_and_meaning_of_quantification_with_Dirk_Geeraerts_and_Dirk_Speelman_).

### 3.0 
Read the data from file to the variable `d_caus`.
```{r 3.0}
d_caus <- read.csv("https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv")
summary(d_caus)
```

### 3.1 
We are going to test whether the association between `Aux` and other categorical variables (`Aux` ~ `CrSem`, `Aux` ~ `CeSem`, etc) is statistically significant. The assiciation with which variable should be analysed using Fisher's Exact Test and not using Pearson's Chi-squared Test? Is this association statistically significant?
```{r 3.1}
fisher.test(d_caus$Aux, d_caus$CeSynt)
```

```
We should use Fisher's Exact Test to test whether the association between `Aux` and `CeSynt` is statistically significant because the sample size is not so large, and there can be frequency values that are less than 5. Due to this condition, it's not possible to test the association using Pearson's Chi-squared Test. P-value is less than 0.05, which means this association statistically significant.
```
### 3.2. 
Test the hypothesis that `Aux` and `EPTrans` are not independent with the help of Pearson's Chi-squared Test. 
```{r 3.2}
test <- chisq.test(d_caus$Aux, d_caus$EPTrans)
test
```

```
P-value is less than 0.05. So, we can consider that `Aux` and `EPTrans` are dependent.
```

### 3.3 
Provide expected frequencies for Pearson's Chi-squared Test of `Aux` and `EPTrans` variables.
```{r 3.3}
test$expected
```
```
          d_caus$EPTrans
d_caus$Aux   Intr     Tr
     doen   40.63  44.37
     laten 198.37 216.63
```
### 3.4. 
Calculate the odds ratio for observed frequencies of `Aux` and `EPTrans`
For 2×2 contigency table
$$\begin{matrix}
a & b \\
c & d
\end{matrix}
$$
one can find *odds ratio* as $(a/c)/(b/d)$.

```{r 3.4}
test$observed
obs_ratio <- (57/182)/(28/233)
obs_ratio
```

### 3.4.1 
Find odds ratio for expected frequencies of `Aux` and `EPTrans`

```{r 3.4.1}
exp_ratio <- (40.63/198.37)/(44.37/216.63)
> exp_ratio
```
What can you say about odds ratio of expected frequencies for arbitrary data?

```
Odds ratio of expected frequencies for arbitrary data is equal to 1. It means that `Aux` and `EPTrans` don't depend on each other.
```

### 3.5 
Calculate effect size for this test using Cramer's V (phi).
```{r 3.5}
library("lsr")
tbl <- table(d_caus$Aux, d_caus$EPTrans)
cramersV(tbl)
```

### 3.6. 
Report the results of independence test using the following template:
```
We have found a significant association between variables `Aux` and `EPTrans` (p < 0.001).  The odds of ratios were 2.6 times higher in group `Intr` than in group `Tr`. Effect size is small (Cramer's V = 0.1691587).
```

### 3.7 
Visualize the distribution using mosaic plot.
Use `mosaic()` function from `vcd` library.
```{r 3.7}
library('vcd')
mosaic(EPTrans ~ Aux, data=d_caus)
```

Below is an example of how to use mosaic() with three variables.
```{r 3.7.1}
# mosaic(~ Aux + CrSem + Country, data=d_caus, shade=TRUE, legend=TRUE)
```

### 3.8 
Why is it not recommended to run multiple Chisq tests of independence on different variables within your dataset whithout adjusting for the multiplicity? (i.e. just testing all the pairs of variables one by one)  
```
The probability of mistakenly rejecting the null hypothesis increases significantly with multiple tests. 
```

### 3.9 
Provide a short text (300 words) describing the hypothesis of this study and the results of your analysis.
```{r 3.9}
The purpose of this study was to analyze an association between the use of two Dutch verbs ('doen' or 'laten') and some other parameters such as the semantic class of the Causer and the Causee, etc. in causative constructions. Moreover, newspaper origin and domains were taken into consideration. Thus, the null hypothesis is the following: an association does not exist, and our variables are independent, while the alternative hypothesis is that a statistically significant association between variables exists. To reject or confirm the null hypothesis, we applied the chi-square test, which is suitable for testing dependence between two categorical variables. The chi-square test on two variables `Aux` and` EPTrans` showed we could reject the H0 and claim that the chosen variables are dependent (p-value was less than 0.05). However, the value of Cramer's V is 0.169, which means small effect size and not so strong association between  `Aux` and` EPTrans.` The other cases of an association between Dutch verbs and different variables should be also analysed with the help of the chi-square test or Fisher's Exact Test,as for an association between `Aux` and `CeSynt.`
```
